Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-06-26 paper/paper.tex

  This paper presents the first Coq formalisation of \emph{hylomorphisms}
  allowing for the mechanisation of all recognised recursive algorithms. The key
  contribution of this paper is that this formalisation is fully
  \emph{axiom-free} allowing for the extraction of safe, idiomatic functional
  code. We exemplify the framework by formalising a series of algorithms based
  on different recursive paradigms such as divide-and conquer, dynamic
  programming, and mutual recursion and demonstrate that the extracted
  functional code for the programs formalised in our framework is efficient,
  humanly readable, and runnable. Furthermore, we show the use of the
  machine-checked proofs of the laws of hylomorphisms to do program
  optimisations.
\end{abstract}

@@

To preserve the generality of hylomorphisms, we encode data structures using
polynomial functors. To ensure that the extracted code is runnable, we make a
conscious effort to stay axiom-free and avoid using indexed types in the
definition of our recursion schemes.

@@

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
A direct encoding of containers as presented in
Section~\ref{sec:category-theory} requires the use of the functional
extensionality axiom, and axiom K to deal with equalities of objects of type
$P(s) \to X$. Axiom K states that a predicate on equality proofs holds for all
such proofs if it holds for reflexivity. This would be needed to reason about
the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
s_2$. In Coq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
$f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
is decidable.  Furthermore, upon extraction Coq will need to insert unsafe casts
for
anything of type $P(s)$.  To avoid these problems, we use \emph{setoids}, and
encode families of positions using \emph{decidable validity predicates}.

@@

As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra $a$ and
$F$-coalgebra $c$, $f$ is a hylomorphism if it satisfies
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, this definition cannot be
accepted by Coq.

In order to find a solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
\emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
the hylomorphism equation. These are hylomorphisms where the coalgebra is
\emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
recursive coalgebras using a predicate that states that a coalgebra terminates
on an input:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
\coq{RecF} represents that a coalgebra will eventually terminate on an input.
The base case takes place when the set of valid positions in the
container extension returned by \coq{h x} is empty.  For convenience, we equip
recursive coalgebras with an additional proof of termination:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
Recursive hylomorphisms are implemented by structural recursion on the proof
that coalgebra \coq{c} eventually terminates for some input \coq{x}
(\coq{RecF c x}).
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
        end (RecF_inv H).
\end{coqcode}

@@

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ => fun _ => nil
  | Node h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In general, we know that if the input is
related to the data inside the structure functor returned by the coalgebra by
some \emph{well-founded relation} \coq{R}, then the coalgebra is recursive. In
this specific case, it is sufficient to show that the two sublists are smaller
than the input list.
\begin{coqcode}
  forall x (p : Pos (shape (qsplit x))), length (cont (qsplit x) p) < length x.
\end{coqcode}
The resulting extracted code is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness~\ref{sec:coq-rec-coalgebras}. The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity.
%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
\footnote{The QSort definition and correctness proof in the
equations
package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
and can be found in the
formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.
\newcommand{\Id}{\text{Id}}
\subsection{Knapsack}\label{sec:knap}
Dynamorphisms are hylomorphisms where the algebra has access to the memory table
of the result of all previous recursive calls. To implement this we need to
formalise the idea of a memory table which is in our case simply represented by
the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
defined as the greatest solution to the following equation
\[
  G_{\infty} A \iso A \times G (G_{\infty} A)
\]
In words, these are the (possibly) infinite trees which branch out with shape
$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
$A \times (1 + G_{\infty} A)$. This is the type representing our memory table
with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
recursive call. Notice that in the case of a recursive $G$-coalgebra this type
only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
$G$-trees thus being able to inspect all the previous recursive calls. This has
been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.

In our framework, we formalise $G$ as a container with shape \coq{Sg} and
position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
$F X = A \times GX$:
\begin{coqcode}
Instance Memo A : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table A := LFix (Memo A).
\end{coqcode}
The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
\begin{coqcode}
Definition consT A : A * App G (Table A) ~> Table A := (* *)
Definition headT A : Table A ~> A := (* *)
Definition tailT A : TableA  ~> App G (Table A) := (* *)
\end{coqcode}
Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
type \coq{Table} post-composed with the \coq{headT} map which outputs the last
result from the memory table:
\begin{coqcode}
Definition dyna A (a : App G (Table A) ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (consT \o pair a id) c.
\end{coqcode}
Note the map \coq{App G (Table A) ~> A} corresponding to a map $G G_{\infty}A \to A$
is not a $G$-algebra.  To recover the $G$-algebra we need to transform it by pairing it with the identity map and
post-composing it with \coq{consT}. This algebra uses this table to lookup
elements, instead of triggering a further recursive call, then elements are
inserted into the memoisation table by the use of \coq{consT} to the result of
applying the algebra. In our example this algebra is called \coq{knapsackA} (see
accompanying code).
% \begin{coqcode}
% Fixpoint memo_knap table wvs :=
%   match wvs with | nil => nil | h :: t =>
%       match lookupT (Nat.pred (fst h)) table with
%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
%       | None => memo_knapsack table t
%       end
%   end.
% Definition knapsack_alg (wvs : list (nat * int))
%   (x : App NatF (Table NatF int)) : int :=
%   match x with | MkCont sx kx => match sx with
%   | inl tt => fun kx => 0%sint63
%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
%   end kx end.
% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
% \end{coqcode}
Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
can define knapsack using the recursion scheme for dynamorphisms:
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
The full extracted code coming from this specification is available in the
artefact accompanying this paper, and an interested reader can check that simple
inlining would produce the following:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note that in the extracted code, the recursive calls to \ocaml{f} build the
memoisation table, and that this memoisation table is used to compute the
intermediate results in \ocaml{memo_knap}, which is finally discarded to produce
the final result.

\subsection{Shortcut Deforestation}\label{sec:shortcut}
We now showcase the shortcut deforestation optimisation on lists.  This is used
to remove the creation of intermediate data structures by fusing multiple
recursive traversals into one.  For example, we can use it to show the following
code from Takano and
Meijer~\cite{TakanoM95}
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
which uses that three different hylomorphisms can be fused into one.
Here \coq{sf1} is the composition of \coq{length},
\coq{Lmap f} and \coq{append ys} which are, in particular,
catamorphisms.

To prove our goal we use the so called \emph{acid rain} theorem which states
that given a parametric function of the following type:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
we can conclude, by parametricity, that the following equation holds:
\begin{coqcode}
  hylo a l_out \o hylo (s l_in) c =e hylo (s a) c
\end{coqcode}
Unfortunately, this is not provable in Coq without adding the parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
functors, e.g. the functor of lists.

@@

There is existing prior work on formalising recursion schemes in a proof
assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted code from
their formalisation.  Mu et al.\ formalise hylomorphisms in Agda, and can do
relational program transformation~\cite{DBLP:journals/jfp/MuKJ09}. It is unclear
how to extract runnable code from their definitions, and they do not seem to
formalise hylomorphisms in terms of generic functors and datatypes.

>>>>

  This paper presents the first Coq formalisation of \emph{hylomorphisms}
  allowing for the mechanisation of all recognised recursive algorithms. The key
  contribution of this paper is that this formalisation is fully
  \emph{axiom-free}, and it allows the extraction of safe, idiomatic functional
  code. We exemplify the framework by formalising a series of algorithms based
  on different recursive paradigms such as divide-and conquer, dynamic
  programming, and mutual recursion and demonstrate that the extracted
  functional code for the programs formalised in our framework is efficient,
  humanly readable, and runnable. Furthermore, we show the use of the
  machine-checked proofs of the laws of hylomorphisms to do program
  optimisations.
\end{abstract}

@@

To preserve the generality of hylomorphisms, we encode data structures using
polynomial functors. To ensure that the extracted code is runnable, avoid using
indexed types in the definition of our recursion schemes. To strengthen the
results of this paper, we avoid all axioms.

@@

\subsection{Mechanising Extractable Container Functors}
\label{sec:containers}
A direct encoding of containers as presented in
Section~\ref{sec:category-theory} requires the use of the functional
extensionality axiom, and axiom K to deal with equalities of objects of type
$P(s) \to X$. Axiom K states that a predicate on equality proofs holds for all
such proofs if it holds for reflexivity. This would be needed to reason about
the equality of any two $f : P(s_1) \to X$ and $g : P(s_2) \to X$, when $s_1 =
s_2$. In Coq, if we know $(s_1 , f) = (s_2 , g)$, we cannot extract a proof that
$f = g$ unless we assume axiom K, or the equality on the type of $s_1$ and $s_2$
is decidable.  Furthermore, upon extraction Coq will need to insert unsafe casts
for
anything of type $P(s)$, because OCaml requires the input value to have a type that is identical to the function parameter, and
in the general case this cannot be done in OCaml.
To avoid these problems, we use \emph{setoids}, and
encode families of positions using \emph{decidable validity predicates}.

@@

As we mentioned in Section~\ref{sec:category-theory}, given an $F$-algebra $a$ and
$F$-coalgebra $c$, $f$ is a hylomorphism if it satisfies
\[
  f = a \circ F\; f \circ c
\]
As we stated earlier, a solution to this equation does not exist for an
arbitrary  algebra/coalgebra pair and, in fact, a recursive function definition
like this cannot be directly accepted by Coq.

In order to find the unique solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
\emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
the hylomorphism equation. These are hylomorphisms where the coalgebra is
\emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
recursive coalgebras using a predicate that states that a coalgebra terminates
on an input:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
\coq{RecF} represents that a coalgebra will eventually terminate on an input.
The base case takes place when the set of valid positions in the
container extension returned by \coq{h x} is empty.  For convenience, we equip
recursive coalgebras with an additional proof of termination:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
Recursive hylomorphisms are implemented by structural recursion on the proof
that coalgebra \coq{c} eventually terminates for some input \coq{x}
(\coq{RecF c x}).
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
        end (RecF_inv H).
\end{coqcode}

@@

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ => fun _ => nil
  | Node h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In general, we know that if the input is
related to the data inside the structure functor returned by the coalgebra by
some \emph{well-founded relation} \coq{R}, then the coalgebra is recursive. In
this specific case, it is sufficient to show that the two sublists are smaller
than the input list.
\begin{coqcode}
  forall x (p : Pos (shape (qsplit x))), length (cont (qsplit x) p) < length x.
\end{coqcode}
We can directly write mergesort as \coq{hylo merge qsplit}, or derive it by
program calculation from the composition of \coq{cata merge}, and the
recursive anamorphism \coq{hylo l_in qsplit}.
The resulting extracted code is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

The proof is by induction on the fact that \coq{qsplit} is recursive. This means
that \coq{qsplit} terminates on \coq{l}.  In practice this is done by using the
tactic \coq{induction (recP qsplit l)}. Then we can unfold \coq{hylo merge qsplit} to
\coq{merge \o fmap (hylo merge qsplit) \o qsplit} using hylomorphism
uniqueness (Section~\ref{sec:coq-rec-coalgebras}). The rest of the proof is standard.
Overall, our proof is shorter than alternative implementations, and of
comparable complexity%
%% Commenting out footnote to save space, since we mention the Coq QSort in related work already...
\footnote{The QSort definition and correctness proof in the
equations
package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.

It would be straightforward to prove the correctness of the fused version, by
simply proving that \coq{Lmap times_two} preserves the order of the elements.
\newcommand{\Id}{\text{Id}}
\subsection{Knapsack}\label{sec:knap}
Dynamorphisms are hylomorphisms where the algebra has access to the memory table
of the result of all previous recursive calls. To implement this we need to
formalise the idea of a memory table which is in our case simply represented by
the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
defined as the greatest solution to the following equation
\[
  G_{\infty} A \iso A \times G (G_{\infty} A)
\]
In words, these are the (possibly) infinite trees which branch out with shape
$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
$A \times (1 + G_{\infty} A)$. This is the type representing our memory table
with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
recursive call. Notice that in the case of a recursive $G$-coalgebra this type
only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
$G$-trees thus being able to inspect all the previous recursive calls. This has
been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.

In our framework, we formalise $G$ as a container with shape \coq{Sg} and
position \coq{Pg}, then \coq{Memo} is the least fixed-point of the functor
$F X = A \times GX$:
\begin{coqcode}
Instance Memo A : Cont (A * Sg) Pg := { valid := valid \o pair (snd \o fst) snd }.
Definition Table A := LFix (Memo A).
\end{coqcode}
The operations of the cofree comonad \coq{Memo} are the \coq{Cons}, the \coq{headT} and the \coq{tailT}:
\begin{coqcode}
Definition consT A : A * App G (Table A) ~> Table A := (* *)
Definition headT A : Table A ~> A := (* *)
Definition tailT A : TableA  ~> App G (Table A) := (* *)
\end{coqcode}
Finally, we define a dynamorphism as a hylomorphism from a type \coq{B} into a
type \coq{Table} post-composed with the \coq{headT} map which outputs the last
result from the memory table:
\begin{coqcode}
Definition dyna A (a : App G (Table A) ~> A) (c : RCoalg G B) : B ~> A
  := headT \o hylo (consT \o pair a id) c.
\end{coqcode}
Note the map \coq{App G (Table A) ~> A} corresponding to a map $G G_{\infty}A \to A$
is not a $G$-algebra.  To recover the $G$-algebra we need to transform it by pairing it with the identity map and
post-composing it with \coq{consT}. This algebra uses this table to lookup
elements, instead of triggering a further recursive call, then elements are
inserted into the memoisation table by the use of \coq{consT} to the result of
applying the algebra. In our example this algebra is called \coq{knapsackA} (see
accompanying code).
% \begin{coqcode}
% Fixpoint memo_knap table wvs :=
%   match wvs with | nil => nil | h :: t =>
%       match lookupT (Nat.pred (fst h)) table with
%       | Some u => (u + snd h)%sint63 :: memo_knapsack table t
%       | None => memo_knapsack table t
%       end
%   end.
% Definition knapsack_alg (wvs : list (nat * int))
%   (x : App NatF (Table NatF int)) : int :=
%   match x with | MkCont sx kx => match sx with
%   | inl tt => fun kx => 0%sint63
%   | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
%   end kx end.
% Definition knapsackA wvs : App NatF (Table NatF int) ~> int :=  (* ... *)
% \end{coqcode}
Finally, given the recursive coalgebra \coq{out_nat} on the natural numbers, we
can define knapsack, given a list of pairs of weights and values \coq{wvs},
using the recursion scheme for dynamorphisms:
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
The full extracted code coming from this specification is available in the
artefact accompanying this paper, and an interested reader can check that simple
inlining would produce the following:
% The reader can check
% in our artefact that the extracted code can be trivially inlined to produce the
% following:
\begin{ocamlcode}
let knapsack wvs x = let (y, _) = (let rec f x0 =
      if x0=0 then
        { lFix_out = {shape = Uint63.of_int (0); cont  = fun _ -> f 0 } }
      else let fn := f (x0-1) in { lFix_out = {
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape in y
\end{ocamlcode}
Note that in the extracted code, the recursive calls to \ocaml{f} build the
memoisation table, and that this memoisation table is used to compute the
intermediate results in \ocaml{memo_knap}, which is finally discarded to produce
the final result.

\subsection{Shortcut Deforestation}\label{sec:shortcut}
We now showcase the shortcut deforestation optimisation on lists.  This is used
to remove the creation of intermediate data structures by fusing multiple
recursive traversals into one.  For example, we can use it to show that the following
code from Takano and
Meijer~\cite{TakanoM95}, which uses three different hylomorphisms, can be fused into one:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here \coq{sf1} is the composition of \coq{length},
\coq{Lmap f} and \coq{append ys} which are, in particular,
catamorphisms.
To prove our goal we use the so called \emph{acid rain} theorem~\cite{TakanoM95}
which states that given a parametric function of the following type:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
we can conclude, by parametricity, that the following equation holds:
\begin{coqcode}
  hylo a l_out \o hylo (s l_in) c =e hylo (s a) c
\end{coqcode}
Unfortunately, this is not provable in Coq without adding the parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}, but we can prove it for specific
functors, e.g. the functor of lists.

@@

There is existing prior work on formalising recursion schemes in a proof
assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted code from
their formalisation.  Mu et al.\ formalise hylomorphisms in Agda, and can do
relational program transformation~\cite{DBLP:journals/jfp/MuKJ09}. Their paper
does not discuss extracting runnable code from their encodings, and they do not
seem to formalise hylomorphisms in terms of generic functors and datatypes.
